Parity games in pushdown systems with parameters

### 2023-07-04 reachability/article.tex
Too long, 15 pages have to shorten to 12

We first consider reachability games: 
the winning condition $\win_F(S)$ is given by a set of final configurations
$F \subseteq S$, and $\win_F(S)$ is the set of maximal plays in $S^* F S^\infty$. 
We simply write
$\win_F$ in case the set $S$ is obvious from context.
To indicate that the winning condition of a game is a reachability winning condition, we will speak of {\em reachability games}.

With regards to reachability games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem.

\problemx{Reachability game}
{A reachability game $\mathcal{G}= ((S_0, S_1,{\rightarrow}),\win_F)$, an initial configuration $s \in S_0 \cup S_1$.}
{Does player $0$ have a winning strategy from $s$ in $\mathcal{G}$? \\}

Secondly, we consider min-parity games: 
the winning condition $\win_\Omega(S)$ is given by a {\em priority function} 
 $\Omega : S \to  [0, m]$.
An infinite path $ \pi = s_0 s_1 s_2 \ldots$ is in $\win_\Omega(S)$ if the smallest priority appearing infinitely often in
the sequence 
 is even, i.e.
if $\min( \{ \Omega(s) | ~ s \in \text{\textit{Inf}}( \pi) \})$ is even.  
A finite play is in $\win_\Omega(S)$ if its target configuration is in $S_1$.
We simply write
$\win_\Omega$ in case the set $S = S_0 \biguplus S_1$ is obvious from context.
To indicate that the winning condition of a game is a parity winning condition, we will speak of
 {\em parity games}.

With regards to parity games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem.

\begin{samepage}
\problemx{Parity game}
{A parity game $\mathcal{G}= ((S_0, S_1,{\rightarrow}),\win_\Omega)$, an initial configuration $s \in S_0 \cup S_1$.}
{Does player $0$ have a winning strategy from $s$ in $\mathcal{G}$?}
\end{samepage}

@@

A winning strategy for player $0$ allows to synthesize a controller that restricts the environment and ensures that {the property expressed by the winning condition always holds.}

@@

@@

For a finite alphabet $\Sigma$,  let $\tau (\Sigma )$ be the vocabulary consisting of a binary relational symbol $\leq$,  and a unary relational symbol $P_a$ for every $a \in \Sigma$. 
A word $w \in \Sigma^*$ can be seen as a $\tau (\Sigma )$-structure where the set of elements consists of a finite interval of $\N$. Elements are seen as positions in the word and for every element $i$, there exists precisely one $a \in \Sigma$ such that $i$ is in the interpretation of $P_a$. Lastly $\leq$ is interpreted as the linear order of $\N$.\\

It is well-known that if we are interested in the complexity of first-order or monadic
second-order model-checking on words, the alphabet can be assumed to be $\{0,1\}$ without loss of generality.
Thus the satisfiability problem is concerned with the following question.
\begin{samepage}
\problemx{$\FO$ satisfiability on words}
{A  
$\FO[\tau(\{0,1\})]$-sentence $\phi$.}
{Does there exists a word $w\in \{0, 1\}^*$, such that $w \models \phi$ ?\newline}
\end{samepage}

@@

\noindent
We start with
 a 
 $\FO[\tau(\{0,1\})]$
-formula $\phi$.
We construct
a parametric pushdown system $\mathcal{Z}_\phi = (Q_\phi,W_\phi, \Delta_\phi )$, with a disjoint
union $Q_\phi = Q_0  \biguplus Q_1 $, a priority mapping $\Omega_\phi$, and an initial state $q_I \in Q_\phi$
such that
player $0$ has a winning strategy from $(q_I, \$, f_I)$ in the  parity game
$\mathcal{G}_{\mathcal{Z_\phi}, Q_0,Q_1, \Omega_\phi}$
if and only if
there exists a word $w \in \{0,1\}^*$ such that
$w  \models \phi $. \\

\noindent
We assume without loss of generality that $\phi$ is written in the prenex normal form
$\phi = \forall x_1 \exists x_2 \forall x_3 \ldots \exists x_k ~ \psi$, where 
an even number $k$ of quantifiers alternate between existential and universal ones, and 
$\psi$ is quantifier-free and in disjunctive normal form.

Player $0$ starts by chosing a valuation $w$ for the first parameter (parameter $p_0$). The goal of this first parameter is to remember
the word $w$, which corresponds to the word ``guessed" by player $0$. 
This parameter is meant in part to enforce that 
the stack remains a prefix of $w$, with each stack content of the play corresponding to a position in the word.
 To enforce this, we allow in $\mathcal{Z}$, after any decision, for a player to challenge the assumption that the last movement of the other player led to a position in the word.
  A player thus challenged only has the ability to add things onto the stack, and win if and only if able to find a way back to the value of parameter $p_0$. 
 The gadgets for these intermediary potential challenges will be excluded from the illustrations to preserve clarity of representation.
The other parameters are meant to correspond to the variables, with player $1$ parameters corresponding to universal variables and player $0$ parameters corresponding to existential variables.

Once players have chosen values for their respective variables, the parameter assignment
$\mu$ is fixed. It is then time
for player $0$ to ensure that
$w$
satisfies $\psi$ if the variables $x_1, \ldots, x_k$ are interpreted by 
$ |\mu(p_1)|-1, \ldots, |\mu(p_k)|-1$ respectively. This is the goal of the reachability game.

The first step of the game is to check that the values assigned to the parameters correspond indeed to possible variables, that is, that the values of the parameters are prefixes of $w$.

Recall $\psi$ is quantifier-free and in disjunctive normal form, i.e.
 \[ \phi = \forall x_1 \exists x_2 \forall x_3 \ldots \exists x_k \bigvee_{i=1}^l(\bigwedge_{j=1}^{h_i} \psi_{i,j}(x_1, \ldots, x_k)).\]

Player $0$, being the existential player, choses one of the conjunctive clause in $\psi$, essentially claiming to be able to prove it. Then player $1$, the universal player, chooses one of the atomic formula in the clause to test. 
Testing the atomic formula is the purpose of a gadget 
in which
 player $0$ has a winning strategy 
  if and only if 
$w$ satisfies 
the atomic formula
 if the variables $x_1, \ldots, x_k$ are interpreted by 
$ |f(1)|-1, \ldots, |f(k)|-1$ respectively.

Now we need only to describe the gadgets $ \C_{\psi_{i,j}}$. 
The gadget depends on the type of the atomic formula. For a formula checking the letter at the position of a variable $x$, player $0$ goes to the position corresponding to $x$, then checks that the top of the stack
$x$ is the right letter. For one checking that $x = x'$ where $x$ and $x'$ are both variables,
player $0$ goes to the position corresponding to $x$ and checks against the parameter corresponding to $x'$ as well. If the formula is a negation, players exchange their roles. 
\end{proof}

>>>>

We first consider {\em reachability games}: 
the winning condition $\win_F(S)$ is given by a set of final configurations
$F \subseteq S$, and $\win_F(S)$ is the set of maximal plays in $S^* F S^\infty$. 
Secondly, we consider {\em parity games}: 
the winning condition $\win_\Omega(S)$ is given by a {\em priority function} 
 $\Omega : S \to  [0, m]$.
An infinite path $ \pi = s_0 s_1 s_2 \ldots$ is in $\win_\Omega(S)$ if the smallest priority appearing infinitely often in
the sequence 
 is even, i.e.
if $\min( \{ \Omega(s) | ~ s \in \text{\textit{Inf}}( \pi) \})$ is even.  
A finite play is in $\win_\Omega(S)$ if its target configuration is in $S_1$.
We simply write
$\win_F$  and $\win_\Omega$ in case the set $S = S_0 \biguplus S_1$ is obvious from context.

@@

@@

Note
{\sc $n$-parametric pushdown parity game} 
is polynomial time reducible to
{\sc $n$-pebble pushdown parity game}.
See appendix for more details.

@@

The satisfiability problem is concerned with the following question.
\begin{samepage}
\problemx{$\FO$ satisfiability on words}
{A  
$\FO$-sentence $\phi$.}
{Does there exists a word $w\in \{0, 1\}^*$, such that $w \models \phi$ ?\newline}
\end{samepage}

@@

\noindent
We start with
 a 
 $\FO$-formula $\phi$.
We assume without loss of generality that $\phi$ is written in the prenex normal form
$\phi = \forall x_1 \exists x_2 \forall x_3 \ldots \exists x_k ~ \psi$, where 
an even number $k$ of quantifiers alternate between existential and universal ones, and 
$\psi$ is quantifier-free and in disjunctive normal form.
We construct
a parametric parity game
in which
player $0$ has a winning strategy
if and only if
there exists a word $w \in \{0,1\}^*$ such that
$w  \models \phi $.

Player $0$ starts by chosing a valuation $w$ for the first parameter (parameter $p_0$). The goal of this first parameter is to remember
the word $w$, which corresponds to the word ``guessed" by player $0$. 
This parameter is meant in part to enforce that 
the stack remains a prefix of $w$, with each stack content of the play corresponding to a position in the word.
 To enforce this, we allow in $\mathcal{Z}$, after any decision, for a player to challenge the assumption that the last movement of the other player led to a position in the word.
  A player thus challenged only has the ability to add things onto the stack, and win if and only if able to find a way back to the value of parameter $p_0$. 
The other parameters are meant to correspond to the variables, with player $1$ parameters corresponding to universal variables and player $0$ parameters corresponding to existential variables.

Once players have chosen values for their respective variables, the parameter assignment
$\mu$ is fixed. It is then time
for player $0$ to ensure that
$w$
satisfies $\psi$ if the variables $x_1, \ldots, x_k$ are interpreted by 
$ |\mu(p_1)|-1, \ldots, |\mu(p_k)|-1$ respectively. 
Recall $\psi$ is quantifier-free and in disjunctive normal form, i.e.
 $ \psi = %\forall x_1 \exists x_2 \forall x_3 \ldots \exists x_k 
 \bigvee_{i=1}^l(\bigwedge_{j=1}^{h_i} \psi_{i,j}(x_1, \ldots, x_k)).$
Player $0$, being the existential player, choses one of the conjunctive clause in $\psi$, essentially claiming to be able to prove it. Then player $1$, the universal player, chooses one of the atomic formula in the clause to test. 
Testing the atomic formula is the purpose of a gadget 
in which
 player $0$ has a winning strategy 
  if and only if 
$w$ satisfies 
the atomic formula
 if the variables $x_1, \ldots, x_k$ are interpreted by 
$ |f(1)|-1, \ldots, |f(k)|-1$ respectively.
The gadget will depend on the type of the atomic formula. For a formula checking the letter at the position of a variable $x$, player $0$ goes to the position corresponding to $x$, then checks that the top of the stack
$x$ is the right letter. For one checking that $x = x'$ where $x$ and $x'$ are both variables,
player $0$ goes to the position corresponding to $x$ and checks against the parameter corresponding to $x'$ as well. If the formula is a negation, players exchange their roles. 
\end{proof}
